---
title: "R Notebook - Project 1 -Beer Analysis"
output:
  pdf_document: default
  html_document: default
---

# Libraries Used and Source Data
```{r Libraries and Source Data}
###Libraries used
library(knitr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(maps)
library(mapproj)
library(usmap)
library(class)
library(caret)
library(e1071)
library(usmap)

###reading in data Beers.csv and Breweries.csv
Beers <- read_csv("Beers.csv")
Breweries <- read_csv("Breweries.csv")

```

# Analysis Q1 - How many breweries are present in each state
```{r Q1}

# Combined data set is grouped sorted by amount of brewries in each state
Breweries %>% 
group_by(State) %>% 
summarize(count = n() )%>% 
arrange(desc(count))%>% 
print(n=51)

## Answer Q1 - The most breweries are Colorado (47), California (39) and Michigan (32). The fewest breweries are in DC, North Dakota, South Dakota and West Viginia (1 each).

# Rename "State" column to "state" for the plot_usmap 
# Group Breweries by state
breweries_gb <- Breweries %>%
  rename(
    state = State
  ) %>%
  group_by(state) %>%
  summarise(
    count=n()
    )

# Merge Beers and Breweries data
full_bar <- merge(Beers, Breweries, by.x = "Brewery_id", by.y = "Brew_ID")

# Rename "State" column to "state" for the plot_usmap 
# Group Breweries by state, get count, abv(&ibu) median and max
full_bar_gb <- full_bar %>%
  rename(
    state = State
  ) %>%
  group_by(state) %>%
  summarize(
    count=n(),
    abv_median = median(ABV, na.rm = TRUE),
    ibu_median = median(IBU, na.rm = TRUE),
    abv_max = max(ABV, na.rm = TRUE),
    ibu_max = max(IBU, na.rm = TRUE)
    )


# Map number of breweries in each State
plot_usmap(data = breweries_gb, values = "count", color = "white") + 
  scale_fill_continuous(
    low = "blue", high = "red", name = "Breweries per State", label = scales::comma
  ) + theme(legend.position = "right")

# Map number of ABV median in each State
plot_usmap(data = full_bar_gb, values = "abv_median", color = "white") + 
  scale_fill_continuous(
    low = "red", high = "green3", name = "Median ABV per State", label = scales::comma
  ) + theme(legend.position = "right")

# Map number of ABV max in each State
plot_usmap(data = full_bar_gb, values = "abv_max", color = "white") + 
  scale_fill_continuous(
    low = "red", high = "green3", name = "Max ABV per State", label = scales::comma
  ) + theme(legend.position = "right")

# Map number of IBU median in each State
plot_usmap(data = full_bar_gb, values = "ibu_median", color = "white") + 
  scale_fill_continuous(
    low = "white", high = "blue", name = "Median IBU per State", label = scales::comma
  ) + theme(legend.position = "right")

# Map number of IBU max in each State
plot_usmap(data = full_bar_gb, values = "ibu_max", color = "white") + 
  scale_fill_continuous(
    low = "white", high = "blue", name = "Max IBU per State", label = scales::comma
  ) + theme(legend.position = "right")

```

# Analysis Q2 - Merging Beweries and Beers data and printing first and last 6 observations
```{r Q2}
#data sets are combined  on common variable Brewery ID
BeersBreweries=merge(Beers,Breweries,by.x = "Brewery_id",by.y = "Brew_ID")
colnames(BeersBreweries)[2]<- "Beer_Name"
colnames(BeersBreweries)[8]<- "Brewery_Name"
head(BeersBreweries,n=6)
tail(BeersBreweries,n=6)
```


# Analysis Q3- Missing values (NA's analysis)
```{r Q3}
#replacing Blank Values with NAs
BeersBreweries [BeersBreweries==""]<- NA

#Finding miss values in the entire data set 
apply(is.na(BeersBreweries),2,sum)

#Filtering on each of the three columns that have missing values
MissingIBU <- BeersBreweries %>% filter(is.na(IBU))
MissingABV <- BeersBreweries %>% filter(is.na(ABV))
MissingStyle <- BeersBreweries %>% filter(is.na(Style))

#Creating "clean"" versions of the data set without missing values
IBUClean <-BeersBreweries %>% filter(!is.na(IBU))
ABVClean <- BeersBreweries %>% filter(!is.na(ABV))

## Answer Q3 - There are 1005 missing IBU values, 62 missing ABV values, and 5 missing Style values. All rows missing the Style were also missing the IBU. We did not make any changes tot he missing values. Due to the large amount of missing values we analyzed a few different solutions such as median, mean or linear relationship but ulitmately decided to remove missing data points for relevant analysis.
```


# Analysis Q4 - Compute median alcohol content and IBU for each state. Bar Chart
```{r Q4}
#Grouping the "clean" IBU data by state, summarizing for median IBU and sorting in order 
IBUCleanSummary<-IBUClean %>% 
group_by(State) %>% 
summarize(medianIBU = median(IBU) )%>% 
arrange(desc(medianIBU))%>% 
print(n=51)

#Bar Chart for the summary above
IBU_barchart <- IBUCleanSummary %>% ggplot() + geom_bar(aes(x=reorder(State, -medianIBU), y=medianIBU), stat="identity", fill="blue") + ggtitle("Median IBU per State") +ylab("Median IBU") + xlab("State") +

  theme(axis.text.x = element_text(angle=90, vjust=.6))

#Grouping the "clean" ABV data by state, summarizing for median ABV and sorting in order 
ABVCleanSummary<-ABVClean %>% 
group_by(State) %>% 
summarize(medianABV = median(ABV))%>% 
arrange(desc(medianABV))%>% 
print(n=51)

#Bar Chart for the summary above
ABV_barchart <- ABVCleanSummary %>% ggplot() + geom_bar(aes(x=reorder(State, -medianABV), y=medianABV), stat="identity", fill="green3") + ggtitle("Median ABV per State") +ylab("Median ABV") + xlab("State") +

  theme(axis.text.x = element_text(angle=90, vjust=.6))

## Answer Q4 - Maine (61 IBU), West Virginia (57.5 IBU), FLorida, and Georgia (both 55 IBU) have the highest median IBU out of all the states. Ariziona (20.5 IBU), Kansas (20 IBU) and Wisconsin (19 IBU) have the lowest median IBUs. DC and Kentucky (both 6.25% ABV)  have the highest median ABV out of all the states. New Jersey (4.6% ABV) and Utha (4% ABV) hav the lowest median ABVs.
```


#####NOT used (plan was to show median ABV and IBU by state on one graph, but were not able to adjust the scaling and sorting properly. Also the graph seemed too clustered.
```{r}
MedianComp=merge(ABVCleanSummary,IBUCleanSummary,by = "State")
MedianComp<- MedianComp[c(1,3,2)]
MedianComp<-MedianComp[order(MedianComp$medianIBU),]
MedianCompMelt <- melt(MedianComp, id.vars = 'State')

ggplot(MedianCompMelt, aes(x= reorder(State, -value), y=value, fill=variable))+
  geom_bar(stat='identity', position='dodge')+
   scale_y_continuous(
    "MedianIBU",
    sec.axis = sec_axis(~ . * .001, name = "MedianABV")
  )
```


# Analysis Q5 - State with max ABV and max IBU
```{r Q5}
#Repeating code from Q4 to find the MAX ABV/IBU per state
IBUCleanSummary2<-IBUClean %>% 
group_by(State) %>% 
summarize(medianIBU = median(IBU) )%>% 
arrange(desc(medianIBU))%>% 
print(n=51)

ABVCleanSummary2<-ABVClean %>% 
group_by(State) %>% 
summarize(medianABV = median(ABV) )%>% 
arrange(desc(medianABV))%>% 
print(n=51)

#Finding the MAX IBU/ABV value and its state for an indivdual brewery.
IBUClean2<-IBUClean[order(-IBUClean$IBU),]
ABVClean2<-ABVClean[order(-ABVClean$ABV),]

## Answer Q5 - The state with the max median IBU is Maine and the states with the max median ABVs are DC and Kentucky. The brewery producing the beer with the highest IBU is located in Astoria Oregon. It is the Astoria Brewing Company and they are producing the "Bitter Bitch Imperial IPA" with an IBU of 138.The brewery producing the beer with the highest ABV is located in Boulder Colorado. It is the Upslope Brewing Company and they are producing the "Lee Hill Series Vol. 5 - Belgian Style Quadrupel Ale" with an ABV of 12.8%.
```


# Analysis Q6 - Summary Statistics and distribution of ABV
```{r Q6}
#Summary Statistics of the combined data set
summary(BeersBreweries)

#Plots for the distribution of ABV and IBU 
ABV_sumstat <- BeersBreweries %>% ggplot() + geom_histogram(aes(x=ABV*100), fill="green3") + ggtitle("Distribution of ABV") + xlab("Percent ABV")

IBU_sumstat <- BeersBreweries %>% ggplot() + geom_histogram(aes(x=IBU), fill="blue") + ggtitle("Distribution of IBU") + xlab("IBU")

 
##Answer Q6 - The beer brewed at the most breweries is the "Nonstop Hef Hop". 75% of beers have an ABV between 5.0% and 6.7%. 75% of beers have an IBU between 21 and 64. The most popular styles of beer are the American IPA, the American Pale Ale and the American Amber/ Red Ale. The brewery breewing the most beers is the "Brewery Vivant" with 62 different beers. The city brewing the most beers is Grand Rapis, brewing 66 different beers. Colorado is the state breewing the most beers with 265 different beers.
```


# Analysis Q7 - Relationship + scatter plot IBU vs. ABV
```{r Q7}
#Combining the two clean sets of data to create one data set with no NAs
IBUABVClean <-ABVClean %>% filter(!is.na(IBU))

#Scatterplot of ABV vs IBU
IBUABVClean %>%
ggplot(mapping = aes(x = ABV, y = IBU)) +
geom_point()+   labs(title="IBU vs ABV correlation", subtitle="Breweries", caption="source: Brewery Data")+geom_smooth(method=lm)

#calculating linear relationship betweem IBU and ABV
fit <- lm(IBU ~ ABV, data = IBUABVClean)
summary(fit)
confint(fit, level = 0.95)

## Answer Q7 - Based on the scatterplot we can visually see a linear positive relationship between ABV and IBU. This is confirmed by running a linear regression analysis on it. The R-squared value is 44.97%, p-value (<2.2e-16) and the 95% confidence interval of the slope is [1207.77, 1356.31].
```


# Analysis Q8 -KNN IBU/ABV Ale vs IPA
```{r Q8 pt.1}
#creating new column in the data set for the "Style Group", which classifies each beer to a style (here IPA vs. Ale) based on text recognition.
IBUABVClean$StyleGroup = NA
IBUABVClean$StyleGroup[grepl("ale",IBUABVClean$Style,ignore.case=TRUE)] = "Ale"
IBUABVClean$StyleGroup[grepl("*ipa",IBUABVClean$Style,ignore.case=TRUE)] = "IPA"

#summarizing data by style group as a check
IBUABVClean %>% 
group_by(StyleGroup) %>% 
summarize(count = n() )

#Scaling the data to prep it for KNN Analysis
IBUABVCleanScaled = data.frame(ZIBU = scale(IBUABVClean$IBU), ZABV = scale(IBUABVClean$ABV), StyleGroup = IBUABVClean$StyleGroup)
IBUABVCleanScaled <-IBUABVCleanScaled %>% filter(!is.na(StyleGroup))

#Running KNN Analysis for k=3, k=7 and k=11
classifications = knn.cv(IBUABVCleanScaled[,1:2], IBUABVCleanScaled$StyleGroup, k = 3)
confusionMatrix(table(classifications,IBUABVCleanScaled$StyleGroup))

classifications2 = knn.cv(IBUABVCleanScaled[,1:2], IBUABVCleanScaled$StyleGroup, k = 7)
confusionMatrix(table(classifications2,IBUABVCleanScaled$StyleGroup))

classifications3 = knn.cv(IBUABVCleanScaled[,1:2], IBUABVCleanScaled$StyleGroup, k = 11)
confusionMatrix(table(classifications3,IBUABVCleanScaled$StyleGroup))

## Answer Q8 - Our KNN model using internal CV predicts the style of beer as either IPA or Ale with ~85% accuracy for k=3, k=7 or k=11. 95% CI is ~ [83.5%, 87.5%] Sensitivity and Specifity of the model are ~88% and ~82 respectively.
```

# Analysis Q8 Part 2
```{r Q8 pt.2}
#Running a naive bayes model with 100 iterations
iterations = 100

masterAcc = matrix(nrow = iterations)

splitPerc = .7 #Training / Test split Percentage

for(j in 1:iterations)
{
  
  trainIndices = sample(1:dim(IBUABVClean)[1],round(splitPerc * dim(IBUABVClean)[1]))
  train = IBUABVCleanScaled[trainIndices,]
  test = IBUABVCleanScaled[-trainIndices,]
  
  model = naiveBayes(train[,c(1,2)],as.factor(train$StyleGroup),laplace = 1)
  table(predict(model,test[,c(1,2)]),as.factor(test$StyleGroup))
  CM = confusionMatrix(table(predict(model,test[,c(1,2)]),as.factor(test$StyleGroup)))
  masterAcc[j] = CM$overall[1]
}
#calculating mean Accuracy of the model
MeanAcc = colMeans(masterAcc)

MeanAcc

## Answer Q8 Part 2 - The Naive Bayes model, on avereage on 100 iterations run, predicts the Style of beer as eiter IPA or Ale with 84.2% accuracy.
```


# Analysis Q9 - Grouping the style of beer based on text recognition
```{r Q9}
#Removing all Style = NA data points from the full data set
StyleClean <- BeersBreweries %>% filter(!is.na(Style)) 

#Unlisting and splitting the Style column of the data set into individual words and sorting them in descending order
theText = unlist(str_split(str_replace_all(StyleClean$Style,"[^[:alnum:] ]", ""), boundary("word")))
sort(table(theText), decreasing=T)

theText2 = unlist(str_split(str_replace_all(StyleClean$Beer_Name,"[^[:alnum:] ]", ""), boundary("word")))
sort(table(theText2), decreasing=T)

#Creating style group labels based on text recognition
StyleClean$StyleGroup = "Miscellaneous"
StyleClean$StyleGroup[grepl("ale",StyleClean$Style,ignore.case=TRUE)] = "Ale"
StyleClean$StyleGroup[grepl("*ipa",StyleClean$Style,ignore.case=TRUE)] = "IPA"
StyleClean$StyleGroup[grepl("lager",StyleClean$Style,ignore.case=TRUE)] = "Lager"
StyleClean$StyleGroup[grepl("stout",StyleClean$Style,ignore.case=TRUE)] = "Stout"
StyleClean$StyleGroup[grepl("porter",StyleClean$Style,ignore.case=TRUE)] = "Porter"
StyleClean$StyleGroup[grepl("pilsener|pilsner",StyleClean$Style,ignore.case=TRUE)] = "Pilsener"
StyleClean$StyleGroup[grepl("beer",StyleClean$Style,ignore.case=TRUE)] = "Other Beer"
StyleClean$StyleGroup[grepl("fruit",StyleClean$Style,ignore.case=TRUE)] = "Fruit"
StyleClean$StyleGroup[grepl("witbier",StyleClean$Style,ignore.case=TRUE)] = "Witbier"
StyleClean$StyleGroup[grepl("kölsch|hefeweizen|märzen|oktoberfest|berliner|weissbier|altbier",StyleClean$Style,ignore.case=TRUE)] = "German Bier"
StyleClean$StyleGroup[grepl("cider",StyleClean$Style,ignore.case=TRUE)] = "Cider"
StyleClean$StyleGroup[grepl("bitter",StyleClean$Style,ignore.case=TRUE)] = "Bitter"

#Summary/count of the above style groups and Pie Chart
StyleCleanSummary<-StyleClean %>% 
group_by(StyleGroup) %>% 
summarize(count = n() )

ggplot(StyleCleanSummary, aes(x="", y=count, fill=StyleGroup)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0) +
  theme(axis.title.x = element_blank(),
  axis.title.y = element_blank(), 
  axis.ticks = element_blank(),
  axis.text = element_blank(),
  panel.grid  = element_blank())


#Another text recogntion step to indetify additional Style groups (used in repitition to add on to the style group steps above until found to be suficciently grouped)
FindMoreTypes <- StyleClean %>% filter(is.na(StyleGroup)) 
theText2 = unlist(str_split(str_replace_all(FindMoreTypes$Style,"[^[:alnum:] ]", ""), boundary("word")))
sort(table(theText2), decreasing=T)

#creating new summary dataframe for further analysis
TotalCount<-StyleClean %>% 
group_by(StyleGroup, State) %>% 
summarize(count = n())

#Calculating Mean ABV and IBU by Style

StyleIBUABVCleanClean <- StyleClean %>% filter(!is.na(IBU))
StyleIBUABVCleanClean <- StyleIBUABVCleanClean %>% filter(!is.na(ABV))

StyleIBUABVCleanSummary<-StyleIBUABVCleanClean %>% 
group_by(StyleGroup) %>% 
summarize(medianABV=median(ABV), medianIBU=median(IBU) )

StyleIBUABVCleanSummary %>%
ggplot(mapping = aes(x = medianABV, y = medianIBU)) +
geom_point()+   labs(title="Median ABV & IBU by Style", subtitle="Breweries", caption="source: Brewery Data")+ geom_text(aes(label=StyleGroup), size=3,hjust = -0.2,
              vjust = 0.1,check_overlap = TRUE,angle=30)

#Scatterplot for all beers colored by type
StyleIBUABVCleanClean %>%ggplot(mapping = aes(x = ABV, y = IBU,fill=StyleGroup, colour= StyleGroup)) +
geom_point()+   labs(title="ABV & IBU by Style", subtitle="Breweries", caption="source: Brewery Data")

#Finding the  Top Style Group beer for each state
TotalCountbyStatePer<-TotalCount%>% group_by(State) %>% mutate(percent = count/sum(count))
TotalCountbyStatePer<-TotalCountbyStatePer[order(TotalCountbyStatePer$State),]
TotalCountbyStatePer%>% print(n = 100)

#Changing the type of the State column to character and removing leading white space
TotalCountbyStatePer$State <-as.character(TotalCountbyStatePer$State)
TotalCountbyStatePer$State  <- gsub('\\s+', '',  TotalCountbyStatePer$State )

#Pie Chart Visualization for the composition of beer style groups by state
p4 <- ggplot(data=TotalCountbyStatePer, aes(x="", y = percent, fill=StyleGroup)) + geom_bar(stat = "identity", position = position_fill()) +
coord_polar(theta = "y") +
    facet_wrap(~ State)  +
  theme(axis.title.x = element_blank(),
  axis.title.y = element_blank(), 
  axis.ticks = element_blank(),
  axis.text = element_blank(),
  panel.grid  = element_blank()) + 
    theme(legend.position='bottom') 
    
p4

#Finding the top (highest percentage) style group for each state
TopTotalCountbyStatePer<-TotalCountbyStatePer%>%group_by(State)%>% filter(percent == max(percent)) %>% filter(1:n() == 1)

#reading in US State map information for plotting
us_states <- map_data("state")
head(us_states)

#creating dataframe with state names and abbreviations to allow the joining of the data
state_abbs <- tibble(state = str_to_lower(state.name), abb = state.abb)

#left joining the data on state names
TopTotalCountbyStatePer2=merge(TopTotalCountbyStatePer,state_abbs,by.x = "State",by.y = "abb")
TotalCountbyStatePerMap <- left_join(us_states, TopTotalCountbyStatePer2, c("region"="state"))

#plotting the top style beer for each state on a US map
p <- ggplot(data = TotalCountbyStatePerMap,
            aes(x = long, y = lat,
                group = group, fill = StyleGroup))

p + geom_polygon(color = "gray90", size = 0.1) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) 
    
# Finding/mapping the states each style is the most popular in 
TotalCountbyStylePer<-TotalCount%>% group_by(StyleGroup) %>% mutate(percent = count/sum(count))
TotalCountbyStylePer<-TotalCountbyStylePer[order(TotalCountbyStylePer$StyleGroup),]
TotalCountbyStylePer%>% print(n = 100)

#Cleaning up state column for leading whitespace and changing type to character
TotalCountbyStylePer$State <-as.character(TotalCountbyStylePer$State)
TotalCountbyStylePer$State  <- gsub('\\s+', '',  TotalCountbyStylePer$State )

#Mapping the Ale production as a percentage of total ale production in the US by state
TotalCountbyStylePerAle <-TotalCountbyStylePer%>%filter (StyleGroup == "Ale")

TotalCountbyStylePerAle=merge(TotalCountbyStylePerAle,state_abbs,by.x = "State",by.y = "abb")
TotalCountbyStylePerAleMap <- left_join(us_states, TotalCountbyStylePerAle, c("region"="state"))

p2 <- ggplot(data = TotalCountbyStylePerAleMap,
            aes(x = long, y = lat,
                group = group, fill = percent))

p2 + geom_polygon(color = "gray90", size = 0.1) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) 
    
#Mapping the IPA production as a percentage of total ale production in the US by state

TotalCountbyStylePerIPA <-TotalCountbyStylePer%>%filter (StyleGroup == "IPA")

TotalCountbyStylePerIPA=merge(TotalCountbyStylePerIPA,state_abbs,by.x = "State",by.y = "abb")
TotalCountbyStylePerIPAMap <- left_join(us_states, TotalCountbyStylePerIPA, c("region"="state"))

p3 <- ggplot(data = TotalCountbyStylePerIPAMap,
            aes(x = long, y = lat,
                group = group, fill = percent))

p3 + geom_polygon(color = "gray90", size = 0.1) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) 
```